---
title: "Real data: drivers, overlays and calibration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Real data: drivers, overlays and calibration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 4, dpi = 120)
```

```{r setup}
library(tidysd)
```

Data enters a model in three ways, all additive to the three layers.

## 1. An exogenous driver

An `input()` is an auxiliary whose values come from data rather than from a
formula. It is declared in the structure like anything else -- so the dependency
graph and the unit check see it -- and only the series itself is a parameter.

```{r bathtub}
carbon_struct <- sd_structure(
  meta(name = "Atmospheric carbon bathtub"),
  stock("ExcessAtmosphericCarbon", units = "MtC", non_negative = TRUE),
  input("emissions", units = "MtC/year"),
  flow("emission_flow",   from = .source, to = "ExcessAtmosphericCarbon",
       units = "MtC/year"),
  flow("natural_removal", from = "ExcessAtmosphericCarbon", to = .sink,
       units = "MtC/year")
)

carbon_eqns <- sd_equations(
  emission_flow   ~ emissions,
  natural_removal ~ ExcessAtmosphericCarbon * removal_constant
)

head(global_emissions_synthetic, 3)

carbon_pars <- sd_parameters(
  constant(removal_constant = 0.01),
  initial(ExcessAtmosphericCarbon = 0),
  input_series("emissions", data = global_emissions_synthetic, interp = "linear")
)

out <- simulate(carbon_struct, carbon_eqns, carbon_pars,
                spec = sim_spec(1751, 2011, dt = 1, time_unit = "year"))
autoplot(out, vars = c("ExcessAtmosphericCarbon", "emission_flow", "natural_removal"))
```

The solver interpolates the series onto its own grid, so the data's row spacing
and the model's `dt` need not agree. `range = "hold"` (the default) holds the
first and last values outside the data range; `"extend"` extrapolates linearly.
A `Date` time column is resolved against `sim_spec(start = )`.

Because the series is a parameter, a counterfactual is a parameter edit -- the
model text never moves:

```{r counterfactual}
flat <- transform(global_emissions_synthetic,
                  emissions = pmin(emissions,
                                   emissions[time == 1990]))
carbon_flat <- update(carbon_pars,
                      input_series("emissions", data = flat, interp = "linear"))
out2 <- simulate(carbon_struct, carbon_eqns, carbon_flat,
                 spec = sim_spec(1751, 2011, dt = 1, time_unit = "year"))
c(actual = max(subset(as.data.frame(out), variable == "ExcessAtmosphericCarbon")$value),
  frozen_at_1990 = max(subset(as.data.frame(out2),
                              variable == "ExcessAtmosphericCarbon")$value))
```

## 2. An observed series, for comparison

`simulate(..., observed = )` stacks the observed columns into the output tibble
tagged `source = "observed"` beside `"model"`, so `autoplot()` overlays them.
Use `map =` when the column names differ from the model's.

```{r observed}
ex <- sd_example("boarding_school_flu")
head(ex$observed, 4)

out <- simulate(ex$structure, ex$equations, ex$parameters, spec = ex$spec,
                observed = ex$observed, map = c(I = "sInfected"))
table(out$source)
autoplot(out, vars = "I")
```

Those are the starting guesses, not a fit -- the model is nowhere near the data
yet.

## 3. Calibration

`calibrate()` takes the same three layers as `simulate()`, plus the observed
frame, a `target` map from model variable to data column, the `free` constants
and their bounds. It returns a fitted `sd_parameters` you pass straight back.

```{r calibrate}
fit <- calibrate(ex$structure, ex$equations, ex$parameters, spec = ex$spec,
                 observed = ex$observed,
                 target = c(I = "sInfected"),
                 free   = c("beta", "infectious_period"),
                 lower  = c(beta = 0,    infectious_period = 1),
                 upper  = c(beta = 0.01, infectious_period = 5))
fit
```

`beta` reproduces the value Duggan's `FME::modFit` reports for this outbreak
(about 2.18e-3 per day). The implied basic reproduction number:

```{r r0}
unname(fit$fit$estimate[["beta"]] * 762 * fit$fit$estimate[["infectious_period"]])
```

And the fitted parameters go back through the same verb:

```{r fitted}
fitted <- simulate(ex$structure, ex$equations, fit$parameters, spec = ex$spec,
                   observed = ex$observed, map = c(I = "sInfected"))
autoplot(fitted, vars = "I")
```

One verb, not a sweep or an optimisation framework. `observed =` on
`simulate()` is independent of the fit -- it just carries the data through to
the output.

## Scenarios

Not data, but the same spirit: many runs, one call. Overrides name constants,
and every run is tagged in a `scenario` column.

```{r scenarios, fig.height = 4.5}
out <- sd_example_run("overshoot")
autoplot(out, vars = c("Capital", "Resource"))
```
