---
title: "Getting started with tidysd"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with tidysd}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 4, dpi = 120)
```

```{r setup}
library(tidysd)
```

## Three layers, one verb

A `tidysd` model is not an object you build up; it is three plain, immutable
values that only meet inside `simulate()`.

```
sd_structure()    what exists, how it is wired      (no math)
sd_equations()    functional forms                  (symbolic)
sd_parameters()   the numbers and data
```

The split is not cosmetic. It is what lets you keep a structure and swap the
functional form, or keep both and swap every number, without editing a line of
the other two.

## The smallest complete model

One stock, one outflow proportional to the gap between the stock and a goal:
a cup of tea cooling toward room temperature.

```{r teacup}
teacup_struct <- sd_structure(
  meta(name = "Teacup cooling"),
  stock("TeacupTemperature", units = "degF"),
  flow("heat_loss_to_room", from = "TeacupTemperature", to = .sink,
       units = "degF/minute")
)

teacup_eqns <- sd_equations(
  heat_loss_to_room ~ (TeacupTemperature - room_temperature) / characteristic_time
)

teacup_pars <- sd_parameters(
  constant(room_temperature = 70, characteristic_time = 10),
  initial(TeacupTemperature = 180)
)

out <- simulate(teacup_struct, teacup_eqns, teacup_pars,
                spec = sim_spec(start = 0, stop = 30, dt = 0.125,
                                method = "euler", time_unit = "minute"))
out
```

Every run returns one long tibble: `time`, `variable`, `value`, `unit`, `type`,
plus a column per subscript dimension and, when you ask for them, `scenario`
and `source`. That is the whole output contract, so `ggplot2` comes free.

```{r teacup-plot}
autoplot(out)
```

## What goes in which layer

|Item                                              | Structure | Equations | Parameters |
|:-------------------------------------------------|:---------:|:---------:|:----------:|
|Stock / flow / aux / lookup / input exists; units  |     x     |           |            |
|Flow `from` / `to` wiring                          |     x     |           |            |
|Flow-rate and auxiliary formulas                   |           |     x     |            |
|Choice of functional form                          |           |     x     |            |
|Derived initial-value form (`init(S) ~ pop - I0`)  |           |     x     |            |
|Constant and stock-initial values                  |           |           |     x      |
|Lookup point data; input series                    |           |           |     x      |

Each stock's initial value comes from exactly one place: a number in
`initial()`, or a formula via `init()` in the equations. Asking for both is an
error, and so is asking for neither.

Numeric literals *are* allowed on an equation right-hand side. A `1 - x/K` or
an `ifelse` threshold is structure, not a parameter.

## Order does not matter

Equations resolve into a dependency graph when the model is bound, so you can
write them in whatever order reads best.

```{r order}
struct <- sd_structure(
  meta(name = "S-shaped growth"),
  stock("Stock", units = "widgets", non_negative = TRUE),
  aux("availability"), aux("effect"), aux("growth_rate"),
  flow("net_flow", from = .source, to = "Stock", units = "widgets/time")
)

# written back to front on purpose
eqns <- sd_equations(
  net_flow     ~ Stock * growth_rate,
  growth_rate  ~ ref_growth_rate * effect,
  effect       ~ availability / ref_availability,
  availability ~ 1 - Stock / capacity
)

pars <- sd_parameters(
  constant(capacity = 10000, ref_growth_rate = 0.10),
  reference(ref_availability = 1),
  initial(Stock = 100)
)

out <- simulate(struct, eqns, pars,
                spec = sim_spec(0, 100, 0.25, time_unit = "time",
                                check_units = FALSE))
autoplot(out, vars = c("Stock", "net_flow"))
```

A genuinely simultaneous loop is reported as one, with the cycle that caused it,
rather than silently resolved.

## Immutability, and variants

Layers never change in place. `update()` returns a new object, which is how you
write a variant.

```{r update}
sir <- sd_example("sir")

# frequency-dependent transmission, as shipped ...
deparse(sir$equations$eqns$lambda$rhs)

# ... swapped for density-dependent, structure and parameters untouched
dd <- update(sir$equations, lambda ~ contact_rate * infectivity * I)
deparse(dd$eqns$lambda$rhs)

# and the original is exactly as it was
deparse(sir$equations$eqns$lambda$rhs)
```

The same applies to parameters:

```{r update-pars}
faster <- update(sir$parameters, constant(recovery_time = 2))
sir$parameters$constants$recovery_time
faster$constants$recovery_time
```

## Checking a model without running it

`sd_validate()` does exactly the binding and validation that `simulate()` does,
and hands back the bound model.

```{r validate}
bound <- sd_validate(sir$structure, sir$equations, sir$parameters, sir$spec)
bound
```

## Units

Units are declared in the structure layer and nowhere else. Constants carry
none, so the only check the grammar can honestly make is the structural one: a
flow's units must be the units of the stock it moves, per unit of simulation
time. Everything unspecified is simply not checked.

```{r units}
bad <- sd_structure(
  stock("Population", units = "people"),
  flow("births", from = .source, to = "Population", units = "people")  # missing /year
)
tryCatch(tidysd:::check_flow_units(bad, "year"), warning = conditionMessage)
```

## Where to go next

* `vignette("builtins")` -- delays, smoothing, and shaping functions.
* `vignette("subscripts")` -- arrayed models over named dimensions.
* `vignette("data")` -- exogenous drivers, observed series and `calibrate()`.
* `vignette("catalogue")` -- all twenty-two worked models.
